---
references:
- id: Velloso2013
  title: Qualitative activity recognition of weight lifting exercises
  author:
  - family: Velloso
    given: E
  - family: Bulling
    given: A
  - family: Gellersen
    given: H
  - family: Ugulino
    given: W
  - family: Fuks
    given: H
  container-title: Proceedings of 4th Augmented Human International Conference
  DOI: 10.1145/2459236.2459256
  publisher: ACM
  page: 116-123
  type: article-journal
  issued:
    year: 2013
---

Week 4 Peer Assignment for Practical Machine Learning
=====================================================

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(caret)
```

# Introduction

This assignment explores data collected from a personal activity monitoring
device. People regularly quantify how much of an activity they do, but
rarely to they qualify how well they do it. The analysis below attempts to
answer that question.

The data for this analysis comes from [Weight Lifting Exercises Dataset](http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har)
[also see @Velloso2013.] and contains data from accelerometers on the
belt, forearm, arm, and dumbbell of six (6) participants. They were asked
to perform barbell lifts correctly and incorrectly in five (5) different
ways. This project will predict how well the participant did the exercise
(represented by the classe variable in the data set) from the accelerometer
data.

# Examine data and preprocess

Load the training and testing data sets and begin exploring.

```{r}
training <- read.csv("pml-training.csv",
                     stringsAsFactors=FALSE)

testing <- read.csv("pml-testing.csv",
                     stringsAsFactors=FALSE)

dim(training)

dim(testing)

str(training)

str(testing)
```

Two things immediately stand out. First, there are a large number of variables
(`r ncol(testing)`). Second, many of the variables have a large number of NAs
and/or a large number of empty strings. Since the two data sets are similar,
the training set will analysed to reduce the number of variables and then the
same preprocessing steps applied to the testing set for testing the models
which predict how well a participant performs the exercise.

The first step is to re-read the training set, making sure that NA character
strings and empty character strings are interpreted as NA.

```{r}
training <- read.csv("pml-training.csv",
                     stringsAsFactors=FALSE,
                     na.strings=c("NA", ""))
```

It was also noted that three (3) of the variables do not contribute to
predicting the quality of movement so they will be removed. They are

Variable       | Reason for removal
-------------- | ------------------
X              | This is just the row number
user_name      | This is the participant's name
cvtd_timestamp | This is not a factor variable nor a numeric variable but an alphanumeric representation on the times specified in raw_timestamp_part_1 and raw_timestamp_part_2

```{r}
xformTraining <- subset(training,
                        select=-c(X, user_name, cvtd_timestamp))

dim(xformTraining)
```

For each column, determine the percentage of NAs in that column and list
the columns where that percentage exceeds 50%.

```{r}
NApct <- apply(xformTraining, 2, function(x) {
    round(sum(is.na(x))/nrow(xformTraining), 2)
  })
NApct[NApct > 0.50]
```

There are `r nrow(NApct)` variables for which the number of NAs exceeds 50%.
In fact, the percentage of NAs in these columns is 98%. A high percentage of
NAs in a predictor provides no information so these variables will be
removed from the data set.

```{r}
xformTraining <- xformTraining[, NApct <= 0.5]

dim(xformTraining)
```

Next, the remaining predictors will be checked for zero- or
near-zero-variance and any predictors which have zero- or
near-zero-variance will be removed.

```{r}
nzv <- nearZeroVar(xformTraining)
if (length(nzv) > 0) {
  xformTraining <- xformTraining[, -nzv]
  dim(xformTraining)
}
```

Check the correlation between the remaining predictors and remove
predictors which are highly correlated.

```{r}
corr <- findCorrelation(cor(xformTraining[,-ncol(xformTraining)]))
xformTraining <- xformTraining[, -corr]

dim(xformTraining)
```


# Bibliography


